Local Spectral Clustering of Density Upper Level Sets

### 2019-01-22 icml/icml_local_density_clustering.tex
added todays changes

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\mathbf{\Phi}}
\newcommand{\Psibf}{\mathbf{\Psi}}

%%% Vectors
\newcommand{\pbf}{\mathbf{p}}
\newcommand{\ebf}[1]{\mathbf{e}_{#1}}
\newcommand{\pibf}{\bm{\pi}}

@@

\theoremstyle{remark}
\newtheorem{remark}{Remark}

@@

\begin{abstract}
	Spectral clustering methods are 
	a family of popular non-parametric clustering tools.
	Recent works have proposed and analyzed 
	\emph{local} spectral methods, which 
	extract clusters using
	locally-biased random walks around a user-specified seed node, and are known to have worst-case guarantees for certain graph-based measures of cluster quality.
	In contrast to existing works, 
	we analyze the personalized PageRank (\ppr) algorithm in the classical statistical learning setup,
	where we obtain samples from an unknown distribution, and aim to identify connected
	regions of high-density (density clusters).
	With respect to a neighborhood graph formed over these samples, we provide guarantees for a pair of cluster quality criteria evaluated on empirical analogues to these density clusters. As a result, the \pprspace algorithm is shown to 
	extract sufficiently salient density clusters, and we 
	analyze the sample-complexity of the algorithm.
\end{abstract}

@@

Let $\mathbf{X} := (x_1, \ldots, x_n)$ with $x_i \in \Rd$ for $i = 1,\ldots,n$. Our statistical learning task is clustering: splitting data into groups which satisfy some notion of within-group similarity and between-group difference.

In particular, spectral clustering methods are a family of powerful non-parametric clustering algorithms. Given a symmetric adjacency matrix $\Abf \in \Reals^{n \times n}$ with $(i,j)$th entry representing the similarity between data points $x_i$ and $x_j$, we form the random walk transition probability matrix $\Wbf$, and corresponding graph Laplacian matrix $\Lbf$\footnote{Often, either of the Laplacian matrices  $\Lbf_{sym} := \Dbf^{\frac{1}{2}}\Wbf \Dbf^{-\frac{1}{2}}$ or $\Lbf_{unn} := \Dbf - \Abf$ are used instead.}:
\begin{equation}
\label{eqn: random_walk_laplacian}
\Wbf := \Dbf^{-1}\Abf; ~~~ \Lbf = \Ibf{n} - \Wbf
\end{equation}
where the degree matrix $\Dbf$ is a diagonal matrix with $\Dbf_{ii} := \sum_j \Abf_{ij}$, and $\Ibf{n}$ is the $n \times n$ identity matrix.

Roughly speaking, spectral clustering techniques first embed the data $\Xbf$ using the spectrum of the graph Laplacian, and subsequently 
use this \emph{spectral embedding} to find a clustering of the data.
When applied to large graphs (or large point clouds) classical global spectral 
methods can be computationally cumbersome and 
insensitive to the local geometry of the distribution of the samples
\citep{mahoney2012,leskovec2010}.
This in turn has led to the investigation of local spectral algorithms \citep{spielman2013,anderson2006,leskovec2010}
which leverage locally-biased spectra computed using random walks around 
a user-specified seed node.

A natural model to consider when analyzing point cloud data such as $\Xbf$ is the following:
\begin{equation}
x_i \sim \Pbb, ~~~ \text{independently, for $i = 1,\ldots,n$},
\end{equation} 
with $f$ the density function of $\Pbb$ with respect to the uniform measure over $\Rd$. In this case, we are interested in understanding what the output of a clustering algorithm on this finite sample reveals about the unknown density $f$. For $\lambda > 0$ and the upper level set $\set{x: f(x) \geq \lambda}$, it is intuitive \cite{hartigan1981,chaudhuri2010} to define clusters as the connected components $\mathbb{C}_f(\lambda)$ of the upper level set; we call these connected regions of high density \textit{density clusters}, and study the ability of spectral methods to identify such clusters.

@@

A somewhat more standard mode of understanding spectral clustering methods is to view them as approximating some graph connectivity criteria.

For $\Abf$ as before, consider the graph $G  = (V,E)$, with vertices $V = \set{v_1, \ldots, v_n}$ corresponding to the $n$ rows of $\Abf$, and (possibly weighted) edges $E = \set{(v_i,v_j,\Abf_{ij}): 1 \leq i < j \leq n, \Abf_{ij} > 0}$ (Since $\Abf$ is symmetric, $G$ is an undirected graph; also, by convention, we preclude self-loops, hence $i \neq j$). There are many \cite{yang2015,fortunato2010} graph-based measures which assess the cluster quality of a subset $S \subseteq V$ (or more generally the quality of a partition $S_1 \cup \ldots \cup S_m = V$, for $m \geq 2$.)

Arguably a natural way to assess cluster quality is via a pair of criteria capturing the \textit{external} and \textit{internal connectivity} of $S$, respectively.  As the names suggest, external connectivity should relate to the number of edges between $S$ and its complement $V / S$ (hereafter denoted $S^c$), while internal connectivity in turn measures the number of edges between subsets within $S$. The graph clustering task then becomes to find a subset $S$ (or, for global algorithms, a partition $S_1 \cup \ldots \cup S_m = V$), which has both small external and large internal connectivity.

We will assess the external connectivity of a subset $S \subseteq V$ through its normalized cut. The cut of $S$ is
\begin{equation*}
\cut(S; G) := \sum_{u \in S} \sum_{v \in S^c} \1\left((u,v) \in E \right)
\end{equation*}
-- where $S^c = V / S$ is the complement of $S$ in $V$ -- and the volume of $S$ is
\begin{equation*}
\vol(S; G) := \sum_{u \in S} \sum_{v \in V} \1\left((u,v) \in E \right).
\end{equation*}
Then, the \emph{normalized cut} of $S$ is given by
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) := \frac{\cut(S; G)}{ \min\left\{\vol(S; G), \vol(S^c; G) \right\} }
\end{equation} 
Intuitively, a set with low \textit{normalized cut} has many more edges which do not cross the cut than edges which do cross the cut.

Given $S \subseteq V$, the subgraph induced by $S$ is given by $G[S] = (S, E_S)$, where $(u,v) \in E_S$ if both $u$ and $v$ are in $S$ and $(u,v) \in E_S$. Letting $\abs{S} = m$, $\Abf_S$ then denotes the $m \times m$ adjacency matrix representation of $G[S]$; similarly, $\Dbf_S$ is the diagonal degree matrix with entries $(\Dbf_{S})_{ii} = \sum_{j: v_j \in S} \Abf_{ij}$, and $\Wbf = \Dbf_S^{-1} \Abf_S$ is the corresponding random walk matrix (again, defined only over $G[S]$.)

Our internal connectivity parameter $\Psi(S)$ will capture the time it takes for the random walk governed by $\Wbf_S$ to mix (that is, approach a stationary distribution) uniformly over $S$. Denoting the stationary distribution of this random walk by $\widetilde{\pibf} = (\widetilde{\pi}_{1}, \ldots, \widetilde{\pi}_{m})$, the \emph{relative pointwise mixing time} $\tau_{\infty}(S; G[S])$ is defined to be: the smallest integer $t_0 > 0$ such that for all $v = v_i,v' = v_j \in S$:
\begin{equation*}
\abs{\frac{e_v \Wbf_S^t - \widetilde{\pi}_{j}}{\widetilde{\pi}_{j}}} \leq \frac{1}{4}
\end{equation*}
for all $t > t_0$
\footnote{Given a starting node $v$ and and a random walk defined by transition probability matrix $\mathbf{P}$, the rotation $e_v \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$ steps.} (Of course, given the definition of $\Wbf_S$ it is well known that the stationary distribution $\pibf_S$ will be defined by $\widetilde{\pi}_{i} = (\Dbf_{S})_{ii} / \vol(S; G[S])$.

Intuitively, the smaller the pointwise mixing time, the more connected every pair of points $v$ and $v'$ are in the graph $G[S]$. Therefore, the internal connectivity parameter $\Psi(S; G)$ is simply one over the mixing time:
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(S; G[S])}
\end{equation}

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no less than $\Psi$, we will refer to it as a $(\Phi,\Psi)$-cluster. Both local \cite{zhu2013} and global \cite{kannan04} spectral algorithms have been shown to output clusters (or partitions) which provably satisfy approximations to the optimal $(\Phi, \Psi)$-cluster (or partition) for a given graph $G$. \footnote{In the case of \cite{kannan04}, the internal connectivity parameter $\phi$ is actually the conductance, i.e. the minimum normalized cut within the subgraph $G[S]$. See Theorem 3.1 for details; however, note that $\phi^2 / \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a lower bound on $\Psi$.}

@@

The random walk matrix $\Wbf$ over the graph $G = (V,E)$ with associated adjacency matrix $\Abf$ is given by (\ref{eqn: random_walk_laplacian}). \pprspace is then defined with respect to the following inputs: a user-specified seed node $v_i \in V$ , and  $\alpha \in [0,1]$ a teleportation parameter. Letting $v = v_i$ for notational simplicity, and $e_{v}$ be the indicator vector for $v$ (meaning $e_v$ has a $1$ in the $i$th location and $0$ everywhere else), the \emph{PPR vector} is given by the recursive formulation
\begin{equation}
\label{eqn: ppr_vector}
\pbf(v,\alpha;G) := \alpha \ebf{v} + (1 - \alpha) \pbf(v,\alpha;G) \Wbf
\end{equation}
We note in passing that, for $\alpha > 0$, the vector $\pbf(v,\alpha;G)$ can be well-approximated by a simple local computation (of a random walk with restarts at the node $v$.) We also point out that, from a density clustering standpoint, since density clusters are inherently local, using the PPR algorithm eases the analysis, and as we will observe in the sequel our analysis requires fewer global regularity conditions relative to more classical global spectral algorithms.

To compute a cluster $\widehat{C} \subset V$ using the \pprspace vector, we will take sweep cuts of $\pbf(v, \alpha; G)$. Denote the entries of $\pbf(v, \alpha; G)$ by $\pbf(v, \alpha; G) = (p_1, \ldots, p_n)$, and let the \emph{stationary distribution} $\pibf$ of the random walk defined by $\Wbf$ be given by
\begin{equation*}
\pibf = (\pi_1, \ldots, \pi_n), ~~~ \pi_j := \frac{\Dbf_{jj}}{\vol(V; G)}.
\end{equation*}
Then, for a number $\beta \geq 0$, the sweep cut $S_\beta$ is
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \set{u_j \in V: p_j > \beta \pi_j}.
\end{equation}
We will choose one such sweep cut to be our cluster estimate $\widehat{C}$.  
(We delay formal introduction of the local clustering algorithm we analyze until \ref{sec: background}, after we have given a method for forming a graph over the data $\Xbf$.)

\paragraph{Large sample behavior.}

Let $(r_n)$ be a sequence of positive numbers. Given a sequence of kernel functions $k_n: \Rd \times \Rd \to [0,\infty)$ of the form $k_n(x,x') = k(\norm{x - x'}/ r_n)$ for $k$ a non-increasing function, and data $\Xbf = \set{x_1, \ldots, x_n}$ sampled from $\Pbb$ as before, form the (weighted, complete) similarity graph $G_n = (\Xbf, E_n)$ with $E_n = \set{k(x_i,x_j): 1 \leq i < j \leq n}$. (Here, $\norm{\cdot}$ is used to denote Euclidean norm).

In this context, continuous analogues to (for instance) normalized cut have been defined, over the data-manifold rather than the graph, and convergence of finite sample graph-theoretic functionals to their continuous counterparts has been shown
\cite{garciatrillos16, arias-castro12, maier11}.
However these continuous analogues are not always easily interpretable -- and their corresponding minimizers not always easily identifiable -- for the particular density function under consideration. Of course, relating these partitions to the arguably more simply defined high density clusters can be also challenging in general. Intuitively, however, under the right conditions such high-density clusters should have more edges within themselves than to the remainder of the graph. We formalize this intuition next.

@@

Our results can now be summarized by the following two points:

\begin{enumerate}
	\item 
	Under a natural set of geometric conditions\footnote{We formally introduce the geometric conditions in Section \ref{sec: background}. They preclude clusters which are too thin and long, or those for which the gap in density between the high density area and the outside is not sufficiently large}, the normalized cut and inverse mixing time of an empirical density cluster $\Cset[\Xbf]$ can be bounded. Theorems \ref{thm: conductance_upper_bound} and \textcolor{red}{Theorem 2} provide an upper and lower bound, respectively

	\item 
	We show these bounds on the graph connectivity criteria have algorithmic consequences for personalized PageRank. An immediate consequence of Theorems \ref{thm: conductance_upper_bound} and \ref{thm: inverse_mixing_time_lower_bound}, along with the previous work of \cite{zhu2013}, is to yield an upper bound on the normalized cut of the set $\widehat{C}$ output by Algorithm \ref{alg: ppr}, as well as upper bounding the symmetric set difference between $\widehat{C}$ and $\Cset[\Xbf]$. Furthermore, in \ref{sec: consistent_cluster_estimation_with_ppr} we show that a careful analysis of the form typical to local clustering algorithms yields Theorem \ref{thm: consistent_recovery_of_density_clusters}, which states that Algorithm \ref{alg: ppr}, properly initialized, performs consistent density cluster estimation in the sense of (\ref{eqn: consistent_density_cluster_recovery}).
\end{enumerate}

@@

\section{Background and Assumptions.}
\label{sec: background}

We begin by introducing an defining well-conditioned density clusters before turning to formally define the PPR algorithm under consideration, Algorithm \ref{alg: ppr}, and discussing the choice of tuning-parameters.

\subsection{Well-conditioned density clusters.}

@@

Let $B(x,r) = \set{y \in \Rd: \norm{y - x} \leq r}$ be a closed ball of radius $r$ around the point $x$.  Given a set $\Aset \subset \Rd$, and a number $\sigma > 0$, define the $\sigma$-expansion of $\Aset$ to be $\Asig = \Aset + B(0,\sigma) = \set{y \in \Rd: \inf_{x \in \Aset} \norm{y - x} \leq \sigma}$. We are now ready to give the assumptions, which we state with respect to a density cluster $\Cset \in \Cbb_f(\lambda)$ for some $\lambda > 0$, and expansion parameter $\sigma > 0$:
\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp: bounded_density}
	\textit{Bounded density within cluster:} There exist numbers $0 < \lambda_{\sigma} <  \Lambda_{\sigma} < \infty$ such that:
	\begin{equation}
	\label{eqn: bounded_density}
	\lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq \Lambda_{\sigma} 
	\end{equation} 
	%and 
	%\begin{equation*}
	%\frac{\diam \Asig}{\sigma} \leq \mu
	%\end{equation*}
	%where $\diam \Asig = \sup \set{d(x,y) : x,y \in \Asig}$

	\item 
	\label{asmp: low_noise_density}
	\textit{Low noise density:} For some $\gamma > 0 $, there exists a constant $c_1 > 0$ such that for all $x \in \Rd$ with $0 < \rho(x, \Csig) \leq \sigma$,
	\begin{equation*}
	\inf_{x' \in \Csig} f(x') - f(x) \geq c_1 \rho(x, \Csig)^{\gamma},
	\end{equation*}
	where $\rho(x,\Aset) = \min_{x_0 \in \Aset} \norm{x - x_0}$ for $\Aset \subset \Rd$.

	\item
	\label{asmp: cluster_separation}
	\textit{Cluster separation:}
	For all $\Cset' \in \Cbb_f(\lambda)$,
	\begin{equation*}
	\rho(\Csig,\Csig') > \sigma,
	\end{equation*}
	where $\rho(\Aset,\Aset') = \min_{x \in \Aset} \rho(x,\Aset')$ for $\Aset' \subset \Rd$.

	\item
	\label{asmp: cluster_diameter}
	\textit{Cluster diameter:}
	There exists $D < \infty$ such that for all $x, x' \in \Csig$:
	\begin{equation*}
	\norm{x - x'} \leq D.
	\end{equation*}
\end{enumerate}

We note that $\sigma$ plays several roles here, precluding arbitrarily narrow clusters and long clusters in \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter}, flat densities around the level set in \ref{asmp: low_noise_density}, and poorly separated clusters in \ref{asmp: cluster_separation}.

>>>>

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\mathbf{\Phi}}
\newcommand{\Psibf}{\mathbf{\Psi}}
\newcommand{\dist}{\mathrm{dist}}

%%% Vectors
\newcommand{\pbf}{\mathbf{p}}
\newcommand{\qbf}{\mathbf{q}}
\newcommand{\ebf}[1]{\mathbf{e}_{#1}}
\newcommand{\pibf}{\bm{\pi}}

@@

\newtheoremstyle{aldenrmrk}
{6pt} % Space above
{6pt} % Space below
{} % Body font
{} % Indent amount
{\itshape} % Theorem head font
{.} % Punctuation after theorem head
{.5em} % Space after theorem head
{} % Theorem head spec (can be left empty, meaning `normal')

\theoremstyle{aldenrmrk}
\newtheorem{remark}{Remark}

@@

\begin{abstract}
	Spectral clustering methods are 
	a family of popular non-parametric clustering tools.
	Recent works have proposed and analyzed 
	\emph{local} spectral methods, which 
	extract clusters using
	locally-biased random walks around a user-specified seed node, and are known to have worst-case guarantees for certain graph-based measures of cluster quality.
	In contrast to existing works, 
	we analyze the personalized PageRank (\ppr) algorithm in the classical statistical learning setup,
	where we obtain samples from an unknown distribution, and aim to identify connected
	regions of high-density (density clusters).
	We introduce a bicriteria for evaluating cluster quality and provide guarantees for these criteria evaluated on empirical analogues to these density clusters. As a result, the \pprspace algorithm run over a neighborhood graph is shown to 
	extract sufficiently salient density clusters.
\end{abstract}

@@

Let $\mathbf{X} := (x_1, \ldots, x_n)$ be a sample drawn i.i.d from a distribution $\Pbb$ with density $f$ supported on $\Rd$. Our statistical learning task is clustering: splitting data into groups which satisfy some notion of within-group similarity and between-group difference.

Spectral clustering methods are a family of powerful non-parametric clustering algorithms. Let $G = (V,E)$ be an unweighted and undirected graph, with $\Abf \in \Reals^{V \times V}$ the symmetric adjacency where $A_{uv} = 1$ if $(u,v) \in G$. We form the random walk transition probability matrix $\Wbf$, and corresponding graph Laplacian matrix $\Lbf$\footnote{Often, either of the Laplacian matrices  $\Lbf_{sym} := \Dbf^{\frac{1}{2}}\Wbf \Dbf^{-\frac{1}{2}}$ or $\Lbf_{unn} := \Dbf - \Abf$ are used instead.}:
\begin{equation}
\label{eqn: random_walk_laplacian}
\Wbf := \Dbf^{-1}\Abf; ~~~ \Lbf = \Ibf{n} - \Wbf
\end{equation}
where the degree matrix $\Dbf$ is a diagonal matrix with $\Dbf_{uu} := \sum_{v \in V} \Abf_{uv}$, and $\Ibf{n}$ is the $n \times n$ identity matrix.

Roughly speaking, spectral clustering techniques first learn an embedding the data $\Xbf$ using the spectrum of the graph Laplacian, and subsequently 
apply a simple clustering technique (such as k-means) to this \emph{spectral embedding}.
When applied to large graphs (or large point clouds) classical global spectral 
methods can be computationally cumbersome and 
insensitive to the local geometry of the distribution of the samples
\citep{mahoney2012,leskovec2010}.
This in turn has led to the investigation of local spectral algorithms \citep{spielman2013,anderson2006,leskovec2010}
which leverage locally-biased spectra computed using random walks around 
a user-specified seed node.

We are interested in understanding what the output of a clustering algorithm on $\Xbf$ reveals about the unknown density $f$. For $\lambda > 0$ and the upper level set $\set{x: f(x) \geq \lambda}$, it is intuitive \cite{hartigan1981,chaudhuri2010} to define clusters as the connected components $\mathbb{C}_f(\lambda)$ of the upper level set; we call these connected regions of high density \textit{density clusters}, and study the ability of spectral methods to identify such clusters.

@@

A somewhat more standard mode of understanding spectral clustering methods is to view them as approximating some graph connectivity criteria. There are many graph-based measures which assess the cluster quality of a subset $S \subseteq V$ (or more generally the quality of a partition $S_1 \cup \ldots \cup S_m = V$, for $m \geq 2$. See \cite{yang2015,fortunato2010} for an overview.)

Arguably a natural way to assess cluster quality is via a pair of criteria capturing the \emph{external} and \emph{internal connectivity} of $S$, respectively.  We define the external connectivity of a subset $S \subseteq V$ to be its normalized cut. For $S' \subset V$, the cut between $S$ and $S'$ is
\begin{equation*}
\cut(S, S'; G) := \sum_{u \in S} \sum_{v \in S'} \1\left((u,v) \in E \right)
\end{equation*}
and the volume of $S$ is
\begin{equation*}
\vol(S; G) := \sum_{u \in S} \sum_{v \in V} \1\left((u,v) \in E \right).
\end{equation*}
Then, the \emph{normalized cut} of $S$ is
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) := \frac{\cut(S; G)}{ \min\left\{\vol(S; G), \vol(S^c; G) \right\} }
\end{equation} 
(where we abbreviate $\cut(S; G) = \cut(S; S^c; G)$).

Given $S \subseteq V$, the subgraph induced by $S$ is given by $G[S] = (S, E_S)$, where $(u,v) \in E_S$ if both $u$ and $v$ are in $S$ and $(u,v) \in E_S$.
Our internal connectivity parameter $\Psi(S)$ will capture the time it takes for the random walk over $G[S]$ to mix (that is, approach a stationary distribution) uniformly over $S$. Letting $\abs{S} = m$, $\Abf_S$ denotes the $m \times m$ adjacency matrix representation of $G[S]$; similarly, $\Dbf_S$ is the diagonal degree matrix with entries $(\Dbf_{S})_{ii} = \sum_{j: v_j \in S} \Abf_{ij}$, and $\Wbf = \Dbf_S^{-1} \Abf_S$ is the corresponding random walk matrix (again, defined only over $G[S]$.)

For $v \in V$ we write
\begin{equation}
\label{eqn: random_walk}
\pibf = (\pi_{u})_{u \in S}, ~~ q_{vu}^{(m)} = e_v\Wbf_S^m e_u
\end{equation}
for the stationary distribution and the $m$-step transition probability of the random walk over $G[S]$ originating at $v$.
\footnote{Given a starting node $v$ and and a random walk defined by transition probability matrix $\mathbf{P}$, the rotation $e_v \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$ steps. }
(Of course, given the definition of $\Wbf_S$ it is of course well known that the stationary distribution $\widetilde{\pibf}$ is given by $\widetilde{\pi}_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$.)

The \emph{relative pointwise mixing time} is defined as 
\begin{equation*}
\tau_{\infty}(\qbf; G[S]) := \min\set{m: \forall u,v \in V, \frac{\abs{q_{vu}^{(m)} - \pi_u}}{\pi_u} \leq 1/4} 
\end{equation*}
where $\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}, \qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$.

The internal connectivity parameter $\Psi(S; G)$ is simply one over the mixing time:
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(\qbf; G[S])}
\end{equation}

The graph clustering task is thus find a subset $S$ (or, for global algorithms, a partition $S_1 \cup \ldots \cup S_m = V$), which has both small external and large internal connectivity. If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no less than $\Psi$, we will refer to it as a $(\Phi,\Psi)$-cluster. Both local \cite{zhu2013} and global \cite{kannan04} spectral algorithms have been shown to output clusters (or partitions) which provably satisfy approximations to the optimal $(\Phi, \Psi)$-cluster (or partition) for a given graph $G$. \footnote{In the case of \cite{kannan04}, the internal connectivity parameter $\phi$ is actually the conductance, i.e. the minimum normalized cut within the subgraph $G[S]$. See Theorem 3.1 for details; however, note that $\phi^2 / \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a lower bound on $\Psi$.}

@@

The random walk matrix $\Wbf$ over the graph $G = (V,E)$ with associated adjacency matrix $\Abf$ is given by (\ref{eqn: random_walk_laplacian}). \pprspace is then defined with respect to the following inputs: a user-specified seed node $v \in V$ , and  $\alpha \in [0,1]$ a teleportation parameter. Letting $e_{v}$ be the indicator vector for $v$ (meaning $e_v$ has a $1$ in the $v$th location and $0$ everywhere else), the \emph{PPR vector} is the solution to the following linear equation:
\begin{equation}
\label{eqn: ppr_vector}
\pbf(v,\alpha;G) := \alpha \ebf{v} + (1 - \alpha) \pbf(v,\alpha;G) \Wbf
\end{equation}
We note in passing that, for $\alpha > 0$, the vector $\pbf(v,\alpha;G)$ can be well-approximated by a simple local computation (of a random walk with restarts at the node $v$.) We also point out that, from a density clustering standpoint, since density clusters are inherently local, using the PPR algorithm eases the analysis, and as we will observe in the sequel our analysis requires fewer global regularity conditions relative to more classical global spectral algorithms.

To compute a cluster $\widehat{C} \subset V$ using the \pprspace vector, we will take sweep cuts of $\pbf(v, \alpha; G)$. To do so we require another parameter $\pi_0$. For a number $\beta \geq 0$, the sweep cut $S_\beta$ is then
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \set{u \in V: p_u > \beta \pi_{0}}.
\end{equation}
We will choose one such sweep cut to be our cluster estimate $\widehat{C}$.

\paragraph{Neighborhood graph.}

To formally introduce the local clustering algorithm we consider, we must first give a method for forming a graph over the data $\Xbf$.

Let $(r_n)$ be a sequence of positive numbers. Given a sequence of kernel functions $k_n: \Rd \times \Rd \to [0,\infty)$ of the form $k_n(x,x') = k(\norm{x - x'}/ r_n)$ for $k$ a non-increasing function, and data $\Xbf = \set{x_1, \ldots, x_n}$ sampled from $\Pbb$ as before, form the \emph{neighborhood graph} $G_n = (\Xbf, E_n)$ with $E_n = \set{k(x_i,x_j): 1 \leq i < j \leq n}$. (Here, $\norm{\cdot}$ is used to denote Euclidean norm).

Algorithm \ref{alg: ppr} will be the simple variant of PPR we analyze. It will take as input the data $\Xbf$ along with user-specified parameters $r, \alpha$, $\pi_0$, and $v \in \Xbf$. 
\begin{algorithm}
	\caption{PPR on a neighborhood graph}
	\label{alg: ppr}	
	{\bfseries Input:} data $\Xbf$, radius $r$, teleportation parameter $\alpha \in [0,1]$, seed node $v \in \Xbf$, target $\pi_0$. \\
	{\bfseries Output:} $\widehat{C} \subset V$.
	\begin{algorithmic}[1]
		\STATE Form the neighborhood graph $G_{n,r}$ as given in (\ref{eqn: neighborhood_graph})
		\STATE Compute PPR vector $\pbf(v, \alpha; G_{n,r})$  as defined by (\ref{eqn: ppr_vector}).
		\STATE For $\beta \in (\frac{3}{10}, \frac{1}{2})$ compute sweep cuts $S_{\beta}$ as defined by (\ref{eqn: sweep_cuts}).
		\STATE Return
		\begin{equation*}
		\label{eqn: sweep_cuts_min_conductance}
		\widehat{C} = \argmin_{\beta \in (\frac{3}{10}, \frac{1}{2})} \Phi(S_{\beta}; G_{n,r})
		\end{equation*}
	\end{algorithmic}
\end{algorithm}

We note that the choice of a specific interval such as $(\frac{3}{10}, \frac{1}{2})$ for the range of $\beta$ is standard in the analysis of \pprspace algorithms.

It is worth calling attention to some other work on computing the normalized cut over neighborhood graphs. In this context, continuous analogues to (for instance) normalized cut have been defined, over the data-manifold rather than the graph, and convergence finite sample graph-theoretic functionals to their continuous counterparts has been shown
\cite{garciatrillos16, arias-castro12, maier11}.
However, in addition to the graph-minimization problem being computationally infeasible, these continuous analogues are not always easily interpretable -- and their corresponding minimizers not always easily identifiable -- for the particular density function under consideration. Of course, relating these partitions to the arguably more simply defined high density clusters can be also challenging in general. Intuitively, however, under the right conditions such high-density clusters should have more edges within themselves than to the remainder of the graph. We formalize this intuition next.

@@

Our results can now be summarized by the following points:

\begin{enumerate}
	\item 
	Under a natural set of geometric conditions
	\footnote{We formally introduce the geometric conditions in Section \ref{sec: background}. They preclude clusters which are too thin and long, or those for which the gap in density between the high density area and the outside is not sufficiently large}, 
	the normalized cut of an empirical density cluster $C[\Xbf]$ can be bounded. Theorem \ref{thm: conductance_upper_bound} provides an upper bound.

	\item Under largely the same set of geometric conditions to Theorem \ref{thm: conductance_upper_bound}, Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} provides a lower bound on the inverse mixing time of a random walk over $C[\Xbf]$. Theorem \ref{thm: inverse_mixing_time_lower_bound} gives a tighter lower bound, but requires the additional (restrictive) assumption of convexity.

	\item 
	In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we show these bounds on the graph connectivity criteria have algorithmic consequences for personalized PageRank. we show that a careful analysis of the form typical to local clustering algorithms yields Theorem \ref{thm: consistent_recovery_of_density_clusters}, which states that Algorithm \ref{alg: ppr}, properly initialized, performs consistent density cluster estimation in the sense of (\ref{eqn: consistent_density_cluster_recovery}).

	\item Corollary \ref{cor: ppr_cluster} follows as an immediate consequence of Theorems \ref{thm: conductance_upper_bound} and \ref{thm: inverse_mixing_time_lower_bound}, along with the previous work of \cite{zhu2013}. It gives an upper bound on the normalized cut of the set $\widehat{C}$ output by Algorithm \ref{alg: ppr}, as well as upper bounding the symmetric set difference between $\widehat{C}$ and $\Cset[\Xbf]$.
\end{enumerate}

@@

\section{Graph Quality Criteria on Well-Conditioned Density Clusters.}
\label{sec: background}

@@

Let $B(x,r) = \set{y \in \Rd: \norm{y - x} \leq r}$ be a closed ball of radius $r$ around the point $x$.  Given a set $\Aset \subset \Rd$, and a number $\sigma > 0$, define the $\sigma$-expansion of $\Aset$ to be $\Asig = \Aset + B(0,\sigma) = \set{y \in \Rd: \inf_{x \in \Aset} \norm{y - x} \leq \sigma}$. We are now ready to give the assumptions, which we state with respect to a density cluster $\Cset \in \Cbb_f(\lambda)$ for some $\lambda > 0$, and expansion parameter $\sigma > 0$:
\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp: cluster_separation}
	\textit{Cluster separation:}
	For all $\Cset' \in \Cbb_f(\lambda)$,
	\begin{equation*}
	\dist(\Csig,\Csig') > \sigma,
	\end{equation*}
	where $\dist(\Aset,\Aset') = \min_{x \in \Aset} \dist(x,\Aset')$ for $\Aset' \subset \Rd$.

	\item
	\label{asmp: cluster_diameter}
	\textit{Cluster diameter:}
	There exists $D < \infty$ such that for all $x, x' \in \Csig$:
	\begin{equation*}
	\norm{x - x'} \leq D.
	\end{equation*}

	\item
	\label{asmp: bounded_density}
	\textit{Bounded density within cluster:} There exist numbers $0 < \lambda_{\sigma} <  \Lambda_{\sigma} < \infty$ such that:
	\begin{equation}
	\label{eqn: bounded_density}
	\lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq \Lambda_{\sigma} 
	\end{equation} 
	%and 
	%\begin{equation*}
	%\frac{\diam \Asig}{\sigma} \leq \mu
	%\end{equation*}
	%where $\diam \Asig = \sup \set{d(x,y) : x,y \in \Asig}$

	\item 
	\label{asmp: low_noise_density}
	\textit{Low noise density:} For some $\gamma > 0 $, there exists a constant $c_1 > 0$ such that for all $x \in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,
	\begin{equation*}
	\inf_{x' \in \Csig} f(x') - f(x) \geq c_1 \dist(x, \Csig)^{\gamma},
	\end{equation*}
	where $\dist(x,\Aset) = \min_{x_0 \in \Aset} \norm{x - x_0}$ for $\Aset \subset \Rd$.
\end{enumerate}

We note that $\sigma$ plays several roles here, precluding arbitrarily narrow clusters and long clusters in \ref{asmp: cluster_diameter} and \ref{asmp: bounded_density}, flat densities around the level set in \ref{asmp: low_noise_density}, and poorly separated clusters in \ref{asmp: cluster_separation}.
